---
title: ""
theme: lumen
output:
      html_document
---

```{r setup, include=FALSE}
library(nlme)
library(tidyverse)
load('models2.Rdata')
knitr::opts_chunk$set()
```


### Autoregressive specification of the error variance: \

The models that we have seen so far assumed independent errors within groups. i.e, the variance-covariance matrix $\Sigma$ or sometimes called $R_i$ of the errors of these models is:

$$\Sigma = \sigma^2I=
\begin{bmatrix}
  1 & 0 & 0 & ... & 0 \\
  0 & 1 & 0 & ... & 0 \\
  . & . & . &     & .\\
  . & . & . &     & .\\
  . & . & . &     & .\\
  0 & 0 & 0 & ... & 1
\end{bmatrix}$$
However, the covariance might be non-zero. There could for example temporal dependence within groups. A way to express this type of dependence is the auto-regressive
correlation model. It uses a single correlation parameter $\rho$ ($-1 < \rho < 1$)and assumes that the
time separation between measurements determines their correlation. In our model, it would be:
$$corr(Y_{i,k,j}, Y_{i,k,j'}) = ρ^{|tj−tj'|}$$
For example, if $\rho = 0.8$ and observations are 10 years
apart, their correlation will be $0.8^{10} = 0.107$. In an auto-regressive model the
correlation will decay as the distance betweeen observations increases. $\rho$ is estimated from the data. The `lmer` function that we had been using cannot handle such specification and so we use `lme` form the `nlme` package.\

This new $\Sigma$ matrix can be expressed as the following, for m time points:
$$\Sigma = \sigma^2
\begin{bmatrix}
  1 & \rho & \rho^2 & ... & \rho^{m-1} \\
  \rho & 1 & \rho & ... & \rho^{m-2} \\
  . & . & . &     & .\\
  . & . & . &     & .\\
  . & . & . &     & .\\
  \rho^{m-1} & \rho^{m-2} & ... & \rho & 1
\end{bmatrix}$$

<br>

#### Example on our Simple Model \

We will try to fit this specification to our simple model.\

```{r}
fit.cor <- lme(fixed = thousands_croot ~ cyear, data = sashhinc,
               random = ~1|hhid,
               correlation = corAR1(form = ~ cyear|hhid),
               method = "ML")
summary(fit.cor)

```
\
The correlation parameter is estimated to be 0 (Phi1). Therefore, we get a diagonal matrix or diagonal $R_i$ with variance $\sigma^2 = 0.74$.\

```{r}
R_i <- getVarCov(fit.cor,type="conditional",individual=1)   # R_i
R_i
```

<br>

#### An other Example! \

The following short example was posted by Asher Strauss [here](https://r-sig-mixed-models.r-project.narkive.com/9ks5NVkW/r-sig-me-estimating-ar1-parameters-of-level-one-error-using-lme).\
\
We use the `Glucose` dataset from `lme`. We want to predict glucose levels using time since last meal of the day as fixed effect and subject as random effect with a random slope for time. We filter for meals taken at 10 AM. \

```{r}
data(Glucose)

fGlucose <- Glucose %>% filter(Meal=="10am")

fit.gluc <- lme(fixed=conc~Time, data=fGlucose,
                random=~1+Time|Subject,
                method="REML",
                na.action="na.omit",
                correlation = corAR1(form=~1+Time|Subject))

summary(fit.gluc)
```
\
We see here that there is indeed some time dependence in the errors (phi=0.43) and so we can visualize the conditional variance-covariance matrix $R_i$ where the covariance decays as we explained above.\

```{r}
R_i <- getVarCov(fit.gluc,type="conditional",individual=1)   # R_i
R_i
```
<br>

#### Other Specifications \

There are a lot of other ways to specify the correlation structure of the errors with the `nlme` package. For example, `corCompSymm` assumes a compounded symmetric structure corresponding to a constant correlation. It means that the covariance in the error does not depend on time between two observations.\

```{r}
fit.cor2 <- lme(fixed = thousands_croot ~ cyear, data = sashhinc,
               random = ~1|hhid,
               correlation = corCompSymm(form = ~ cyear|hhid),
               method = "ML")
summary(fit.cor2)
```

```{r}
R_i <- getVarCov(fit.cor2,type="conditional",individual=1)   # R_i
R_i
```
\
We see that the correlation is really small so it is probably fair to assume that the errors are actually independent in this data.\
\
More information on different specifications can be found [here](https://www.rdocumentation.org/packages/nlme/versions/3.1-152/topics/corClasses)

