---
title: "CHNS -- Longitudinal Example"
theme: lumen
output:
      html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(patchwork)
library(haven)
library(lattice)
sashhinc=read_sas("hhinc_10.sas7bdat")
knitr::opts_chunk$set()
```


```{r Packages, eval=FALSE}
library(tidyverse)
library(lme4)
library(patchwork)
require(haven)
library(lattice)
sashhinc=read_sas("hhinc_10.sas7bdat")
```


For this example, we use a subset of the China Health and Nutrition Study (CHNS) Household Income dataset.\
\

### Simple Longitudinal Analysis \
\

We want to understand the evolution of household income over time and so we will start with a very simple model. 
Of course, since we have repeated measures for the same households, there is correlation in our data that we cannot ignore. A linear mixed model handles this naturally by grouping measurements by households. It uses a random intercept for each household.  We use the `lmer` function from the `lme4` package for this analysis in a frequentist framework\

Before we start the modeling effort, we need to do some data preprocessiong and Exploratory Data Analysis. For interpretability, we shift the `year` variable to code the beginning of the study (1989) as 0. We also recode the response variable household income per capita to be expressed in thousands of 2015 Yuan.  \

```{r Data trans}
# let's make this more interpretable by letting 1989 be year 0
# also let's convert income to thousands of yuan
# recode province so we know what they are

sashhinc = sashhinc %>% mutate(
    cyear = WAVE-1989,
    thousands = hhinc_cpi/1000,
    thousands_pc = hhincpc_cpi/1000,
    lnthousands = log(thousands -min(thousands)+1),
    Prov_2 = as.factor(case_when(
      t1==21~'Liaoning',
      t1==11~'Beijing',
      t1==23~'Heilongjiang',
      t1==31~'Shanghai',
      t1==32~'Jiangsu',
      t1==33~'Zhejiang',
      t1==37~'Shandong',
      t1==41~'Henan',
      t1==42~'Hubei',
      t1==43~'Hunan',
      t1==45~'Guangxi',
      t1==52~'Guizhou',
      t1==53~'Yunnan',
      t1==55~'Chongqing',
      t1==61~'Shanxi')))
```

```{r include=FALSE}
sashhinc <- sashhinc %>% mutate(
  area_urban = as.factor(urban))
```

\
We start by looking at the response variable: household income per capita adjusted in 2015 Yuan using CPI, an inflation measure. We see that the raw data is not quite normally distributed. A log transformation does not improve the shape of the distribution and is not convenient as we are dealing with some negative values. A good alternative is a cube root that is shown on the third plot. It allows to deal with negative values and correct some of the issues with the raw data.\

```{r, fig.width=4, fig.height=3}
histogram(sashhinc$thousands, breaks = 600, xlim = c(-30, 300),
          xlab="Imcome (Thousands of Yuan)")
histogram(log(sashhinc$thousands-min(sashhinc$thousands, na.rm=T)+1),
          xlim = c(6.4,7),breaks = 600,
          xlab = "log(Income+1-min(Income)) (Thousands of Yuan)")
histogram(sign(sashhinc$thousands) * abs(sashhinc$thousands)^(1/3), breaks = 50,
          xlim=c(-5, 10),xlab = "Cube Root of Income (Thousands of Yuan)")
```
\
We then want to see how income evolves over time and look at household trajectories.It is quite hard to actually see the different households as there are so many. However, we see that the trend over time is positive meaning that real income per capita increases and household are getting richer overall.\

```{r message=FALSE, warning=FALSE}
sashhinc <- sashhinc %>% mutate(
    thousands_croot = sign(thousands) * abs(thousands)^(1/3))

p1 <- ggplot(data=sashhinc, aes(cyear, thousands_croot)) +
        geom_point(alpha=.1) +
        ylim(-5,12.5) +
        geom_smooth() +
        xlab("Years Since 1989") +
        ylab("Cube Root of Income (thousands of yuan)") + 
        ggtitle("CHNS Household Income by Year",
                "Overall Trend")

p2 <- ggplot(sashhinc, aes(x=cyear,y=thousands_croot,group=hhid)) +
        ylim(-10,13) + 
        geom_point(alpha=0.05) + geom_line(alpha=0.1) +
        #geom_abline(data=est, aes(intercept=int, slope=sl),col="red") +
        xlab("Years Since 1989") +
        ylab("Cube Root of Income (thousands of yuan)") + 
        ggtitle("CHNS Household Income by Year",
                "Household trajectories")

p1 + p2
```
\

It seems like there is a slight upward trend over time overall. We decide to fit a simple linear mixed model with household as random intercept and the cube root of income per capita in 2015 Yuan that we will refer to as income per capita for simplicity.\

```{r message=FALSE, warning=FALSE}
fit.null=lmer(thousands_croot ~ (1|hhid), data=sashhinc) 
fit.1=lmer(thousands_croot ~ cyear + (1|hhid), data=sashhinc)
anova(fit.null, fit.1)
```
\
Including a linear term for time actually improves the model. We would also like to know if the trend differs by household and so we try to fit a random slope.\

```{r}
fit.2=lmer(thousands_croot~cyear + (1+cyear|hhid),data = sashhinc)
```
\
What is the "singular" warning. It is due to the fact that we only have one observation for some households. As a result, we cannot estimate any slope for these households.\
\
We could also test for the linearity of the relationship between income and WAVE by fitting polynomials.\

```{r Time linear, message=FALSE, warning=FALSE}
fit.3 = lmer(thousands_croot ~ poly(cyear, 2) + (1|hhid), data = sashhinc)
fit.4 = lmer(thousands_croot ~ poly(cyear, 3) + (1|hhid), data = sashhinc)
anova(fit.1, fit.3, fit.4)
```
\
A second degree polynomial might be appropriate. However, in order to keep the example simple, we will proceed assuming a linear relationship between time and income per capita.\

\
When looking at the diagnostics plot, the residuals do not follow the normal distribution. The residual Vs. fitted plot has a funnel shape indicating heteroskedasticity.\

```{r}
par(mfrow=c(1,2))
plot(fitted(fit.1), resid(fit.1), xlab = 'fitted', ylab='residual', main='fitted vs. residual')
abline(h=0)
qqnorm(resid(fit.1))
qqline(resid(fit.1))
```

\
Despite these bad diagnostics, we can still interpret the results.\

```{r}
sum1 <- summary(fit.1)
knitr::kable(sum1$coefficients, caption = "Summary of the original model")
```
\
The model estimates that on average, household income per capita in 1989 was  $2.078^3=8.97$ thousands Yuans (2015).
Every 10 years, household income par capita was expected to increase by  $(10*0.05073)^3=0.13$ thousands Yuans (2015). 
<br>
\
Bellow, choose what you prefer to visualize. In the plot including all the points, the red line shows the predicted household income per capita for each year for a new household. In The plot including only one household, the green line shows the predicted household income per capita for this household. Click on "Go!" to view a different household. The household displayed have more than 5 observations.\  
<br>




